miniKanren with fair search strategies

### 2019-03-29 Paper/paper.tex
more intro

\title{miniKanren with fair search}
\author{Kuang-Chen Lu}
\affiliation{Indiana University}
\author{Weixi Ma}
\affiliation{Indiana University}
\author{Daniel P. Friedman}
\affiliation{Indiana University}

@@

@@

The syntax of a programming language should reflect its semantics. When using a
disjunction operator in relational programming, a programmer would expect all
clauses of this disjunction to share the same chance of being explored, as
these clauses are written in parallel. The existing multiarity disjunctive
operator in miniKanren, however, prioritize its clauses by the order of which
these clauses are written down. We have devised two new search strategies that
allocate computational effort fairly in all clauses.

(TODO: evaluation of performance)

@@

With interleaving depth-first search, disjunctive operator allocates half
computational effort to its first goal, and quater to the second, eighth to the
third, and so on. If a goal that can generate planty of unwanted partial 
solutions is placed at the beginning of a \texttt{conde} expressions, it might
waste a lot computational resource. Consequently, miniKanren programmers 
sometimes need to order their \texttt{conde} clauses carefully to prevent some 
clauses absorbing unproportional computational effort. For example, when a 
relational proof checker is used as a proof generator, the eliminational rules 
should comes later.
%TODO KC: prefer an example from other literatures
Seasoned miniKanreners usually know how to utilize the unfairness to optimize
their programs. However, we believe search strategies less sensitive to goal
order can also be useful to little miniKanreners as well as seansoned ones. We 
propose two such search strategies, balanced interleaving DFS (biDFS) and
breadth-first search (BFS), and observe how they affect the efficiency and
query result of known miniKanren programs. The experiment is conducted with the
miniKanren from \textit{The Reasoned Schemer, 2nd Edition}.

>>>>

\title{miniKanren with fair search strategies}
\author{Kuang-Chen Lu}
\affiliation{Indiana University}
\author{Weixi Ma}
\affiliation{Indiana University}
\author{Daniel P. Friedman}
\affiliation{Indiana University}

@@

% Forget the issues about efficiency
% Read the monad paper

@@

The syntax of a programming language should reflect its semantics. When using a
disjunction operator in relational programming, a programmer would expect all
clauses of this disjunction to share the same chance of being explored, as
these clauses are written in parallel. The existing multiarity disjunctive
operator in miniKanren, however, prioritize its clauses by the order of which
these clauses are written down. We have devised two new search strategies that
% allocate computational effort fairly in all clauses.
allocate computational effort more fairly in all clauses.

@@

When every sub-goal of a disjunction produces infinite states, the existing 
disjunctive operator allocates half computational effort to its first goal, and 
quater to the second, eighth to the third, and so on. The unfairness provides 
both opportunity and burden: miniKanren programmers can place more productive 
goal at the beginning to optimize their programs; however, it might be a 
catastrophe if a necessary goal is placed at the end. Seasoned miniKanreners 
usually know how to utilize the unfairness to optimize their programs. However, 
we believe search strategies less sensitive to goal order can also be useful to 
little miniKanreners as well as seansoned ones. We propose two such search 
strategies, balanced interleaving DFS (biDFS) and breadth-first search (BFS), 
and observe how they affect the efficiency and query result of known miniKanren 
programs. The experiment is conducted with the miniKanren from \textit{The 
Reasoned Schemer, 2nd Edition}.

% ---from here to the end of this section are materials for this section---

% Compare with Backtracking, Interleaving, and Terminating Monad Transformers
The \texttt{interleaving} operator has trouble with \texttt{nevero}. Its 
\texttt{conde} suffers from unfairness, but biDFS and BFS does not. This paper 
defines \textit{unfairness} as: "a non-deterministic choice between two 
alternatives tries every solution from the first alternative before any 
solution from the second alternative".

However, like Seres's work, this search cannot handle infinite failure.

% Compare with The Algebra of Logic Programming by Seresall
Seres's work is in Haskell, hence having an easy implementation of stream:
List. Our work is in Scheme. So we have to be explicit about lazyness.
Her definition of fairness is: "a fair search strategy would share the 
computation effort more evenly between the two branches of the computation of 
\texttt{||}, and a fair selection rule would allow one to chose the
literals in a different order."

Her interleaving DFS is the same as the interleaving in the monad paper, but 
they are different from what's known as iDFS in miniKanren community.

Her implementation of fair disjunction cannot handle nevero (or "infinite 
failure" in her terminology): "Note that this implementation of operators does 
not give breadth-first search of the search tree; it deals with infinite 
success but not with infinite failure. Even in the interleaved implementation, 
the first element of the answer list has to be computed before we can switch 
branches; if this takes an infinite number of steps the other branch will never
be reached."

And her streams are all infinitely long, hence querying all answers doesn't 
terminate.

% Observation:
Stream cannot model \texttt{nevero}... The system knows a stream by nevero has no end, but doesn't know the stream will never produce an item.

If a goal that can generate planty of unwanted partial 
solutions is placed at the beginning of a \texttt{conde} expressions, it might
waste a lot computational effort. Consequently, miniKanren programmers 
sometimes need to order their \texttt{conde} clauses carefully to prevent some 
clauses absorbing unproportional computational effort. For example, when a 
relational proof checker is used as a proof generator, the eliminational rules 
should comes later.
%TODO KC: prefer an example from other literatures
